Two subtle problems with over-representation analysis

### 2024-06-06 manuscript/manuscript.Rmd
ms

---
title: "Importance of using the correct background list for over-representation analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

Alternate title: *Problems in over-representation analysis software*

[Mark Ziemann<sup>1</sup>*](https://orcid.org/0000-0002-7688-6974),
[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352)

@@

1. Deakin University, Geelong, Australia, School of Life and Environmental Sciences.

2. Vellore Institute of Technology, Vellore, India.

(*) Corresponding author: m.ziemann@deakin.edu.au

## Abstract

* Over-representation analysis (ORA) is the most common form of functional enrichment and one of the
most used techniques in computational biology, yet there are common methodological issues that
undermine its validity.

* These problems are typically ascribed to errors in analysis and reporting by the end-users,
however software design is another contributor.

* Here, we identify and characterise a potential bug popular ORA software
package and web tool.

* We show that this bug deleteriously affects the results of ORA tests, and we present a simple
work-around to circumvent it.

@@

library("DiagrammeR")

@@

Functional enrichment analysis is a family of computational approaches used to interpret large omics
data.
The most commonly used method is over-representation analysis (ORA) which involves the selection of
genes of interest, followed by a statistical test to evaluate whether genes of certain annotated
categories are over-represented in the set of interest.
The set of interest may be defined in different ways, such as genes that meet a differential expression
significance threshold, or those that have been clustered into a group based on co-expression patterns.
There are various sources of annotated categories, including gene ontologies, protein-coding features,
gene co-expression and signalling or metabolic pathways.
Providers of such annotations include but are not limited to the gene ontology consortium, Reactome,
KEGG, WikiPathways and MSigDB.

Central to ORA tests is the selection of a list of background genes.
This is also sometimes called the "universe" or the "reference" list.
These are defined as any gene that is detected robustly, and has a chance at making the set of interest.
For RNA-seq studies, this could be a threshold detection limit based either on raw counts or counts per
million.

The first use of ORA-based functional enrichment that we are aware of is a 1999 report that used the
hypergeometric test to determine the chance probability of observing over-representation of certain
annotated functions in co-regulated genes in yeast [@Tavazoie1999-kf].
This general approach is the basis of most ORA-based significance testing and can be represented by the
following equation:

$$
P=1-\sum_{i=0}^{k-1}\frac{ {f \choose i} {g-f \choose n-1} }{g \choose n}
$$

Where (P) is the probability of observing at least (k) genes from a functional category within the
list of selected genes with size (n).
Here, (g) is the total number of detected genes, and (f) is the number of detected genes belonging to a
functional category.

In a hypothetical example, there were 134 genes classified as detected.
Of these, 38 were defined as upregulated in response to a stimulus based on a differential expression
test.
From the 38, 8 genes belonged to the "cytokine signaling pathway" category.
This category contains 26 genes in total, but only 18 were detected in this study.

So the background can be defined as 18/134 and the foreground as 8/38 and the enrichment score
expressed as (8/38)/(18/135)=1.58

In this example:

*k*=8

*n*=26

*g*=134

*f*=18

We can calculate the probability of getting exactly 7 members of the set.

$$
P=\frac{ {18 \choose 7} {134-18 \choose 26-1} }{134 \choose 26}
$$

$$
P=\frac{ 480700 * 1.71e29 }{ 3.84e+33 } = 1/21.4 = 0.047
$$

This needs to be repeated for other values of *i*, namely 6 to 0 and then sum all of these.

```{r,hyperdist}
res<-unlist(lapply(0:38, function(i) {  (choose(18,i) * choose((134-18),(38-1)) / choose(134,38)) }))
plot(res,type="b")
```

>>>>

---
title: "Two subtle problems with over-representation analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

[Anusuiya Bora<sup>1,2</sup>](https://orcid.org/0009-0006-2908-1352),
[Mark Ziemann<sup>1,2</sup>*](https://orcid.org/0000-0002-7688-6974)

@@

1. Burnet Institute, Melbourne, Australia.

2. Deakin University, School of Life and Environmental Sciences, Geelong, Australia.

(*) Corresponding author: mark.ziemann@burnet.edu.au

# Abstract

TODO.

# Keywords

* Bioinformatics

* Reproducibility

* Enrichment analysis

* Pathway analysis

* Gene set enrichment analysis

* Docker

# Main Body

@@

library("DiagrammeR")
library("RColorBrewer")
library("kableExtra")
#library("gt")

@@

Functional enrichment analysis involves the summarisation of omics data to reflect
biological changes, and has become one of the most popular methods in bioinformatics.
Enrichment analysis software packages have been collectively cited 131,332 times as of
late 2019 [@Xie2021-ch].
The algorithms used by these tools can largely be divided into two groups, functional
class scoring (FCS) and over-representation analysis (ORA).
In FCS, all detected genes are ranked by their degree of differential expression followed
by some statistical test for enrichment of gene categories at either extreme of the
ranked list.
ORA involdes the selection of genes of interest, followed by a test to ascertain whether
certain functional categories are over-represented in the selected genes.

Of these two methods, FCS is reported to be the most accurate, while ORA seems to be the
most popular.

According to a study that used simulated differential expression data, FCS accuracy was
superior to ORA over a range of experimental designs [@Kaspi2020-rn].

A survey of articles published in 2019 showed that seven of the top eight most popular
methods were based on ORA.

The general approach for over-representation analysis was developed in 1999 and

According to a survey of related publications in 2022,

is one of the most frequently used methods in
computational biology and involves the summarisation of omics data to reflect biological changes,
such as the detection of pathway activation in development or disease
[@Slonim2002-sf; @Khatri2012-do].
As of August 2020, the various types of enrichment analysis were collectively cited 131,332 times [@Xie2021-ch].

The most popular tool for enrichment analysis appears to be DAVID [@Sherman2022-hf].
Its 11 related journal articles have received a total of 37,035 citations from 27,129
separate PubMed articles (as of 29th June 2023).
Examination of related keywords in PubMed suggests the number of articles describing
"enrichment analysis" is doubling every two years, as omics analysis becomes accessible to
more researchers around the world.

These observations highlight that pathway enrichment tools are essential in the interpretation of omics data.
However, there are concerns that inadvertent misuse of these methods leads to invalid results
[@Timmons2015-ex; @Reimand2019-uw; @Wijesooriya2022-li; @Zhao2023-ok].
According to a recent survey of 186 articles, incorrect methods are very common [@Wijesooriya2022-li].
For example, a suitable background list for over-representation tests is reported in only 4.1% of studies.
Correction of p-values to account for multiple testing was reported in only 54% of studies.
This survey also revealed a general lack of reporting detail around other aspects of enrichment analysis,
such as the tool version used.

We hypothesise that the quality of enrichment analysis studies could be dramatically improved by adopting
established simple rules for reproducible computing [@Sandve2013-lk], and specific guidelines for
reporting enrichment analysis [@Zhao2023-ok].

Our aim is to determine how reproducible published enrichment analyses are.
To achieve this, we systematically examined the reproducibility of 20 published articles that used a web-based
enrichment tool.
We investigate root causes of reproduction failure and describe some potential solutions.

## A systematic assessment of reproducibility in enrichment analysis

### Methods

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID (Database for Annotation,
Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
The basis for these selection criteria is that this combination represents the most common application
of enrichment analysis according to our previous study, which could be reproduced without the need to
reanalyse raw data. These articles are listed in the bibliography
[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].

We screened reporting criteria of these articles, including whether the version of the application
is given, whether p-values underwent correction for multiple testing, whether a background list was
described and whether gene selection was described unambiguously.

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring [@Zeeberg2004-na,Ziemann2016-pv, Abeysooriya2021-uz].
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID v6.8, as at this time v6.7 was not available
(conducted March-May 2022) [@Sherman2022-hf].

Briefly, we used the DAVID v6.8 "Analysis Wizard" and pasted in the provided gene list into the text box,
selecting this as "gene list", together with the appropriate gene identifier type.
Typically, this was the official gene symbol.
The species was specified as *Homo sapiens*.
Next, we clicked on the "Functional Annotation Tool" link, bringing up the "Annotation Summary Results"
page.
We followed the steps indicated in the respective papers to select appropriate gene sets investigated
in the articles such as Gene Ontology (GO; biological process, molecular function), Kyoto Encyclopedia
of Genes and Genomes (KEGG) pathways, etc.
Next, the "Functional Annotation Chart" was clicked on, opening a new window with a table of
enrichment results, which were saved for future reference.

The pathway enrichment results presented in the paper were compared to those we obtained.
It would be ideal to compare standardised effect sizes (fold enrichment scores) between original and
reproduction results, however these were reported in the main text of only 3/20 articles.
Therefore we adopted a binary classification system, based on head-counting pathways that were presented as
statistically significant in the original article.
We defined a pathway as confirmed when it reached a similar level of statistical significance
upon reproduction.
We define "similar significance level" to be a value within 50% of the value presented in the original article.
For example if an article shows a pathway with FDR=0.04, then an acceptable FDR in reproduction could be in
the range of 0.02 to 0.06.
Additionally, statements made in the abstract, results, and discussion sections based on specific pathway
enrichment results were carefully examined and cross-checked with our results to understand whether those
assertions survived reproduction or not.

Each article was critically evaluated and given a score of one (1; low) to three (3; high).
Articles were rated 1 when <50% of pathways shown in the results were confirmed
upon reproduction, and most conclusions derived from enrichment analysis were not supported by reproduction.
Articles were rated 2 when ≥50% of pathways shown in the results were confirmed
upon reproduction, and most conclusions derived from the pathway enrichment analysis were supported.
Articles were rated 3 when ≥80% of terms shown in the original article were observed again in
reproduction with a similar significance value, and all conclusions derived from the pathway analysis were
supported in reproduction.

### Results

As reproducibility is closely associated with the quality of documentation, we collected key methodological
information about how the enrichment analysis was done and reported (Table 1).
Only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.

```{r,table1,echo=FALSE}

x <- read.table("../data/checklist.tsv",header=TRUE,row.names=1)

colnames(x) <- gsub("_"," ",colnames(x))
x[,3]=NULL
x |> gt(rownames_to_stub = TRUE) |> tab_header(title = "Table 1. Article reporting checklist results.")
#x %>% kbl(caption = "Table 1. Article reporting checklist results.") %>% kable_paper("hover", full_width = F)

We attempted reproduction according to the methods described in the respective articles, with the limitation
that we had to use DAVID v6.8 as v6.7 was no longer available.
The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
Seven studies were classified as low reproducibility, while nine were classified as medium and just four
were classified as high.
Justifications for these classifications are provided in Table S1.

```{r,barchart,fig.width=6,fig.height=5,echo=FALSE,fig.cap = "Figure 1. Examining reproducibility of enrichment analysis in a sample of 20 journal articles. Beside the PubMed Central identifier, the reported version of DAVID used in the original publication (blank if none given)."}

score1 <- c("PMC6405693 (v6.7)","PMC6425008 (v6.8)","PMC6535219","PMC6539328 (v6.8)",
  "PMC6542760","PMC6561911 (v6.8)","PMC6663624")
score2 <- c("PMC6381667","PMC6463127","PMC6557785","PMC6580941","PMC6591946",
  "PMC6582306","PMC6333352 (v6.8)","PMC6526186","PMC6607402 (v6.7)")
score3 <- c("PMC6349697 (v6.8)","PMC6444048","PMC6587650", "PMC6368841")
l <- list("1-low"=score1,"2-medium"=score2,"3-high"=score3)

barplot(unlist(lapply(l,length)),col="lightgray",
  xlab="Reproducibility score",ylab="no. articles",
  ylim=c(0,11))

#grid()
text(0.7,1:7-0.5,labels=score1,cex=0.85)
text(1.9,1:9-0.5,labels=score2,cex=0.85)
text(3.1,1:4-0.5,labels=score3,cex=0.85)

pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
pc2 <- paste(pc,"%",sep="")
text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)

pdf("fig1.pdf",width=6,height=5)
barplot(unlist(lapply(l,length)),col="lightgray",
  xlab="reproducibility score",ylab="no. articles",
  ylim=c(0,11))
grid()
text(0.7,1:7-0.5,labels=score1)
text(1.9,1:9-0.5,labels=score2)
text(3.1,1:4-0.5,labels=score3)
pc <- unlist(lapply(l,length)) / sum(unlist(lapply(l,length))) * 100
pc2 <- paste(pc,"%",sep="")
text(c(0.7,1.9,3.1),unlist(lapply(l,length))+0.4,labels=pc2)
dev.off()

| S.No. Pubmed ID: Replication of (Table/Figure) | A. Was Replication possible? B. Were there other significant terms found which were not mentioned in the article? | Replication score and justification |
| --- | --- | --- |
| 1.PMC6349697: Figure 2C and 2D | A. Yes, however, corrected p values are different after replication. B. Yes, only BP terms were taken into account so several CC and MF significant terms were not mentioned in the paper. |  **3 (high).** Considering top 6 BP Terms and KEGG pathways. The basis of the selection of just 6 gene ontologies and 4 KEGG pathways shown in the paper needs to be addressed, when there is a higher amount of significant GOs and pathways that are observed after replication. It was impossible to determine if the article utilised a background gene list or not. The steps to perform DAVID analysis could have been elaborated in the paper. The results shown in the paper are observed after replication but only 6 BP GO significant terms are shown in the paper. |
| 2.PMC6368841: Figure 3 A - D | A. Only GO Terms were the similar as the paper. The FDR values are inconsistent. However, the paper should have clearly mentioned the different tools under figure captions to bring more clarity of the terms obtained by respective tools like DAVID and ClusterProfiler. The KEGG terms do not come up in DAVID at all. B. No | **3 (high).** The article receives a score of 3 solely based on GO results. GO results from the original manuscript are somewhat similar to reproduced analysis in DAVID. The FDR values however are inconsistent. The list of genes as input into DAVID analysis is unclear. The paper should have mentioned that both up and down-regulated proteins were considered for DAVID analysis. Another important question is why were two different tools used for GO and KEGG when both can be performed in DAVID or clusterprofiler? |
| 3.PMC6381667: Figure 1 and Figure 2 | A. Results are not reproducible and are highly inconsistent B. N.A. | **2 (medium).** GO terms obtained after replication showed inconsistent results compared to the results in the original article. KEGG results were similar. Better naming convention for supplementary file for 117 genes for DAVID input is necessary. GO Results are not reproducible and are highly inconsistent. It was not made clear which results were obtained with DAVID or IPA. |
| 4.PMC6405693: Table 1 and Table 2; Texts under “Protein-Protein Interaction Network Building and Interrelation Analysis Between Pathways” | A. Results do not match. B. N.A. | **1 (low).** GO analysis from up-regulated and down-regulated proteins are less similar to the results given in the paper. The FDR values for some pathways are highly inconsistent. KEGG Analysis of all DEGs do not match at all. |
| 5.PMC6425008: Most of the tables and figures of the paper (Figure 5, 6, 7), Table 2 | A. No. B. N.A. | **1 (low).** This paper could have done proper labelling of supplementary files as it was difficult to find the gene list for DAVID analysis. Lacks information on when control genes are used and when citral-affected gene lists are used. Replication of GO and KEGG analysis is not consistent with results from the paper. How were Figures 5-7 derived? The overall protocol is not so coherent. |
| 6.PMC6444048: The text under “Pathway analysis of SUVmax related genes” And Table 2 | A. Terms are the same but gene counts have been different. The DAVID results are somewhat reproducible. B. No | **3 (high).** Gene counts in the terms significantly enriched in the “process of cell division” and “nucleoplasm” have changed. Only cell cycle is highly enriched in the replicated results, the claims of metabolic pathways and hypoxia signalling (through GSEA analysis) to be enriched are not accurate. The results claimed are somewhat similar. All the processes have different FDR values. |
| 7.PMC6463127: Table 4, Table 5, Table 6 | A. Yes but the significance differed. B. Yes, several terms from replication were not mentioned in the paper for GO Analysis. | **2 (medium).** Does not consider corrected p values. Overall, significant terms from GO analysis of up and down-regulated DEGs are missing upon replication. |
| 8.PMC6535219: Figure 2 | A. Somewhat similar. B. There were many top significant terms that were not mentioned. | **1 (low).** Narrowing/Filtering the given gene list is a required step that could have been mentioned. Many terms were not so significant anymore. There were new terms that are of top positions after replication. |
| 9.PMC6539328: Figure 1A | A. Mostly not. B. Yes, many | **1 (low).** The significant pathways stated in the paper are not showing up once replicated. There is no clarity of the up-regulated genes used for GO and KEGG analysis. Only 21 up-regulated genes are listed whereas the paper states 4831. Also, why are only up-regulated genes used? |
| 10.PMC6542760: Figure 4 | A. No. B. Yes, many significant terms were observed. | **1 (low).** Replication results of GO terms are not at all similar with the results shown in the paper. |
| 11.PMC6557785: Figure 5 A-B | A. Yes, but results in paper are not significant upon replication. B. Yes | **2 (medium).** For GO terms, there are other terms that have higher numbers of genes, making them more enriched than immune-related pathways as mentioned in the paper. For KEGG Pathways, there are other pathways that are more significant and contain a higher number of genes enriched for example “Salmonella infection” that has a higher count of genes but is not mentioned in the paper. |
| 12.PMC6561911: Figure 5 a and b | A. No. B. N.A. | **1 (low).** The enrichment analysis results are inconsistent with the results displayed in the paper. There was no KEGG analysis done |
| 13.PMC6580941: Table 1 and Table 2 | A. Yes. B. Yes! Significant terms like “nucleoplasm”, “RNA Binding”, “protein binding” are missing from the paper. | **2 (medium).** For GO analysis, the replicated results are not as significant as the GO terms shown in Tables 1 and 2 of the paper. The Gene list for input is not so clearly mentioned, it should have been labelled properly in the name of the supplemental file. P-values are completely different. |
| 14.PMC6587650: Figure 2 | A. Yes. B. Yes, some. | **3 (high).** (Best) The GO terms results of BP, MF, CC and KEGG pathway results from the paper are compliant with replicated results. They could have mentioned why they chose the type of sorting shown in the paper to display their results. Overall, all results have been replicated. |
| 15.PMC6591946: Figure 5 c and Figure 6 a | A. Yes for GO analysis only. B. Yes for KEGG analysis | **2 (medium).** The GO term results of replication are compliant with Figure 5 c of the paper. KEGG pathway results are semi-consistent. |
| 16.PMC6663624: Table 3 and Table 4 | A. Yes. B. Yes | **1 (low).** The top 2 significant GO terms in the paper are not significant after replication analysis is done. Other GO terms are the same as replicated results. For KEGG Analysis, some terms are more significant after replication. Eg: PD-L1 expression and PD-1 checkpoint pathway in cancer, Th1 and Th2 cell differentiation, Th17 cell differentiation were more significant after “natural killer (NK) cell-mediated cytotoxicity” KEGG Pathway. The paper is somewhat replicable. |
| 17.PMC6582306: Figure 5 | A. Partly. B. Yes, many. | **2 (medium).** The paper uses 44 sets of DEGs grouped in 4 major types. For replication, OFC Set 9 was used, DAVID GO analysis was done twice where a discrepancy of results was observed. Compared to the paper, the most significant term was missing after replication analysis was conducted. For OFC set 9, several pathways are seen after replication but only one pathway is shown in paper. This should be addressed. |
| 18.PMC6333352: Figure 7 | A. Yes. B. Yes. There was a difference in gene counts as well. | **2 (medium).** DAVID was used for KEGG Analysis: From WGCNA - Bahn - KEGG Analysis: Epstein-Barr virus infection, the pathway with the highest gene count is missing in the results shown in the paper. From NERI - BAHN - KEGG Analysis: “Pathways of neurodegeneration - multiple diseases”, “Pathways in cancer” - pathways with the highest gene count (30 and 28 respectively) are not mentioned in the paper. Other pathways are present and the paper is somewhat reproducible. |
| 19.PMC6526186: Table 1 | A. Yes. B. Yes. | **2 (medium).** The top BP GO terms from the paper are no longer significantly higher in the replicated GO results. The top 2 KEGG Pathways after replication are “Human cytomegalovirus infection” followed by “Pathways in cancer”. The p-values have changed after the results were replicated, compared to the values given in the paper. The rest of the pathways are similar to the results given in the paper. |
| 20.PMC6607402: Figure 1 A-D | A. Yes. B. Yes, many! | **2 (medium).** For GO analysis and their replication, top enriched BP, CC and MF GO terms are not mentioned. Most of the terms are mentioned in the replicated results, but different levels of enrichment. |

: Table S1. Classification and justification of reproducibility scores for each article.

The ability to reproduce DAVID enrichment analysis using the supplementary gene
lists was highly variable.
The version of DAVID used is likely to be a big contributor to irreproducibility as we know that there are major
differences in the underlying annotation set (knowledgebase) used by version 6.7 and 6.8.
This is supported by the fact that none of the v6.7 studies received a high reproduicibility score.
On the other hand these versions did not receive regular updates, so the knowledgebases were static for relatively
long times; v6.7 (Jan. 2010 - Oct 2020) and v6.8 (Oct 2016 - Jun 2022).
We can then deduce that knowledgebase version is likely not the cause of irreproducibility for PMC6561911,
PMC6539328 and PMC6425008; studies that reportedly used v6.8 but received low reproducibility scores.

We note that DAVID version 6.8 was retired in June 2022, so these 20 studies are currently no longer reproducible
with the original tools.
The implications of this are outlined in the Discussion section below.

## A five-pillars approach to addressing reproducibility of enrichment analysis

Our results emphasise the need for better methodological reporting to reduce ambiguity in methods sections
and highlight that webtools can suddenly become unavailable.
Although there are various solutions to these issues (refer to the Discussion section), here we will outline
a potential solution whose design is guided by current best practices
[@Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Ziemann2023-dg].
Our aim is to provide a beginner-friendly guide to enrichment analysis that will be
reproducible for many years to come.

### Method overview

We have developed a step-by-step protocol for how to conduct a highly reproducible
enrichment analysis, which is attached as a supplementary file (Suppl Info 1) and published to
protocols.io with a permissive license [@Ziemann2023-kz].
This protocol is prepared in accordance with the *five pillars* principles [@Ziemann2023-dg], and includes
version control with git and GitHub, compute environment control using Docker, literate programming
with R Markdown, persistent code and data sharing with Zenodo and documentation of all of these to
empower others to reproduce the workflow.
Central to this protocol is the provision of template Dockerfile and an R Markdown script, which
are intended to be remixed and altered by end users to suit their project's needs.
The general steps involved are shown in **Figure 2**.

```{r,method_diagram,echo=FALSE,fig.height=6,fig.cap = "Figure 2. Flow chart of example workflow. An estimated amount of time to conduct each step is given."}

library("RColorBrewer")
library("shape")

cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))

polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)

polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)

polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)

polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)

polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
text( 6.3, 2, lab="5. Customise R Markdown: 3-12 hrs",cex=1.5)

polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)

polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)

polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)

#Arrows(2, 7.5, 2.5, 7.0, lwd=2, arr.type="triangle")
#Arrows(2.5, 6, 3, 5.5, lwd=2, arr.type="triangle")
#Arrows(3, 4.5, 3.5, 4, lwd=2, arr.type="triangle")
#Arrows(3.5, 3, 4, 2.5, lwd=2, arr.type="triangle")
#Arrows(4, 1.5, 4.5, 1, lwd=2, arr.type="triangle")

pdf("fig2.pdf",width=8,height=6)
cols <- brewer.pal(5, "Set2")
cols2 <- adjustcolor( cols, alpha.f = 0.7)
plot(0,type='n',axes=FALSE,ann=FALSE, xaxs="i", yaxs="i", xlim=c(0, 11), ylim=c(-3, 8.5))

polygon(c(0,6,6,0),c(8.5,8.5,7.5,7.5), col=cols2[1], border = NA)
text( 3, 8.0, lab="1. Install Docker: 7 mins",cex=1.5)

polygon(c(0.5,7.5,7.5,0.5),c(7.0,7.0,6.0,6.0), col=cols2[2], border = NA)
text( 3.9, 6.5, lab="2. Validate example: 14 mins",cex=1.5)

polygon(c(1,8.5,8.5,1),c(5.5,5.5,4.5,4.5), col=cols2[3], border = NA)
text( 4.8, 5.0, lab="3. Create GitHub repo: 29 mins",cex=1.5)

polygon(c(1.5,10.5,10.5,1.5),c(4,4,3,3), col=cols2[4], border = NA)
text( 6.1, 3.5, lab="4. Build custom Docker image: 45 mins",cex=1.5)

polygon(c(2,10.5,10.5,2),c(2.5,2.5,1.5,1.5), col=cols2[5], border = NA)
text( 6.3, 2, lab="5. Customise R Markdown: 3-12 hrs",cex=1.5)

polygon(c(2.5,9.5,9.5,2.5),c(1,1,0,0),col="lightgray", border = NA)
text( 6.1, 0.5, lab="6. Update repository: 70 mins",cex=1.5)

polygon(c(3,11,11,3),c(-0.5,-0.5,-1.5,-1.5), col=cols2[1], border = NA)
text( 7.0, -1, lab="7. Update Docker image: 40 mins",cex=1.5)

polygon(c(3.5,12,12,3.5),c(-2,-2,-3,-3), col=cols2[2], border = NA)
text( 7.6, -2.5, lab="8. Long Term Archiving: 180 mins",cex=1.5)
dev.off()

The Dockerfile uses a base image from Bioconductor that corresponds to the latest available
stable release version, on the Ubuntu operating system.
The image contains instructions for installing a few useful utilities including `nano` and `git` for
modifying scripts, and `magic-wormhole` for transferring data between computers.
It installs R packages from CRAN and Bioconductor repositories to fulfill common tasks in transcriptome
analysis such as fetching RNA-seq counts from DEE2 with `getDEE2` @Ziemann2019-zo,
differential expression with `DESeq2` @Love2014-hd,
enrichment analysis with `clusterProfiler`, `fgsea` and `mitch`
[@Yu2012-ts; @Korotkevich2016-ls; @Kaspi2020-dq].
It also contains instructions to clone a GitHub repository, which contains the scripts that run the
analytical workflow.
Lastly, it copies a gene set library from Reactome @Gillespie2022-gz with a clear version
history to guarantee provenance of the annotation data.

The R Markdown script called `example.Rmd` undertakes the transcriptome and enrichment analyses.
R Markdown allows the unification of extended descriptions, code and results into a single document,
which is becoming ubiquitous in bioinformatics and data science more generally
@Grolemund2017-zl.
R Markdown supports languages other than R, including Python, Shell, Julia and others.
The ability to provide extended descriptions is helpful for providing contextualising information about
the purpose of the work, detailed methods and interpretations derived from the analysis.
The R Markdown script contains code chunks that perform specific tasks including fetching an example
data set from DEE2, inspecting data quality, differential expression and two types of enrichment
analysis: over-representation analysis (ORA) and functional class scoring (FCS) (**Figure 3**).

```{r,diagram2,echo=FALSE,fig.cap = "Figure 3. Flow chart of example workflow."}

library("DiagrammeR")
grViz("digraph flowchart {
      node [fontname = Helvetica, shape = rectangle]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
      tab3 -> tab5 -> tab6 }
      [1]: 'Fetch RNA-seq from DEE2'
      [2]: 'Quality control'
      [3]: 'Differential expression'
      [4]: 'Over-representation analysis'
      [5]: 'Functional class scoring'
      [6]: 'Euler diagram'
")

grViz("digraph flowchart {
      node [fontname = Helvetica, shape = rectangle]
      tab1 [label = '@@1']
      tab2 [label = '@@2']
      tab3 [label = '@@3']
      tab4 [label = '@@4']
      tab5 [label = '@@5']
      tab6 [label = '@@6']
      tab1 -> tab2 -> tab3 -> tab4 -> tab6;
      tab3 -> tab5 -> tab6 }
      [1]: 'Fetch RNA-seq from DEE2'
      [2]: 'Quality control'
      [3]: 'Differential expression'
      [4]: 'Over-representation analysis'
      [5]: 'Functional class scoring'
      [6]: 'Euler diagram'
") %>%
  DiagrammeRsvg::export_svg() %>%
  charToRaw() %>%
  rsvg::rsvg_pdf("fig3.pdf")

For each enrichment analysis conducted, a checklist is provided to comply with reporting guidelines
[@Wijesooriya2021-lb; @Zhao2023-ok].
An example checklist is shown in **Table 2**.

| Reporting criteria | Method or resource used |
| -- | -- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | `clusterProfiler` |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via dee2.io at accession SRP038101 |
| Other parameters | Minimum gene set size of 10 |

: Table 2. An example reporting checklist for enrichment analysis.

The software versions used in this example analysis are provided in Table 3.

| Software | Version | Purpose |
| --- | --- | --- |
| Ubuntu | 22.04 | Operating System |
| R | 4.3.0 | Statistical computing language |
| nano | 6.2 | text editing |
| git | 2.34.1 | source control |
| R/kableExtra | 1.3.4 | render formatted tables |
| R/vioplot | 0.4.0 | violin charts |
| R/gplots | 3.1.3 | heatmaps |
| R/eulerr | 7.0.0 | Venn diagrams |
| BioC/getDEE2 | 1.6.0 | fetch example RNA-seq data |
| BioC/DESeq2 | 1.36.0 | differential expression |
| BioC/fgsea | 1.22.0 | functional class scoring |
| BioC/clusterProfiler | 4.4.4 | over-representation analysis |
| BioC/mitch | 1.10.0 | multi-contrast enrichment |

: Table 3. The software bundled in the example Docker image.

Reproducing the example workflow is relatively quick and simple, requiring six commands and approximately 4-7
minutes starting from a base Linux install.
The steps include installing Docker, pulling and running the container, running the R Markdown script,
exiting the container and copy the newly created report from the container to the current working
directory (Box 1).

```bash
sudo apt update && sudo apt install docker.io -y  # 30 sec
sudo docker run --platform linux/amd64 mziemann/enrichment_recipe
Rscript -e 'rmarkdown::render("example.Rmd")' # 150 sec
docker cp $(docker ps -aql):/enrichment_recipe/example.html . # instant
firefox example.html
```
<p style="text-align: center;">
*Box 1. Linux shell commands to reproduce a transcriptome and enrichment analysis using a Docker image.
Ubuntu 22.04 was used.*
</p>

### Demonstration

To demonstrate the utility of this workflow, we present the results of this example data analysis;
it is a comparison of control and 5-azacitidine-treated AML3 cell transcriptomes with RNA-seq
in triplicate @Lund2014-et.
We use this analysis to better understand the effect of DNA methyltransferase inhibitors on
gene expression and elaborate upon previous findings [@Lund2014-et], which mainly focused on
the relationship between RNA expression and DNA methylation but did not show results of a
transcriptome-only enrichment analysis.
Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
FCS, and give guidance for interpreting results of such tests.

For quality control the number of reads per sample was calculated and found to be in the range of
11-16M.
Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
genes.
A multi-dimensional scaling plot showed clustering of samples into their treatment
groups, suggesting a robust effect of the chemical exposure (**Figure 4A**).
Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
(**Figure 4B**).
Up and down-regulated gene sets were subjected to ORA separately, using the list of
13,168 genes as the background, and gene sets from REACTOME.
